\chapter{Introduction}

Context-sensitive techniques, data fusion and ranking agents are frequently used in industry applications and consumer software products. Combined, these techniques are useful for ranking social media according to a users preferences and context. This report describes an agent for the classifying, scoring and ranking of data according to its context-sensitive relevance to a user.

\section{The Problem}

Social media, productivity tools and internet-based information are abundant on mobile devices, leading to users being overwhelmed with information, despite only a small amount of it being of any interest to a particular individual at any given moment. This calls for a means by which such data can be ranked or filtered according to its importance, interest or relevance.

\section{Project Objective}

The objective of this project is to produce a scalable and highly modular context-sensitive mobile-content relevance-based intelligent ranking agent, to order social media, productivity and other web-based information according to its time- and situation-specific relevance to a user.

\section{Goals}

The following are core goals which this project sets out to achieve.

\begin{enumerate}
	\item Develop a scoring algorithm by which to judge to relevance of an item of data based upon a users
	\begin{enumerate}
		\item Personality profile
		\item Historical data such as click-history
		\item Environment (conditional upon time-constraints)
	\end{enumerate}
	\item To perform automatic remote topic analysis to judge the topic of an item of data
	\item Develop a sorting/ranking algorithm to sort or insert scored items of data efficiently, into an ordered list
	\item Develop a stable and robust data fusion technique to combine a range of data into a user-context and data-context object.
	\item Abstract away this agent into an extensible Java API for use in
	\begin{enumerate}
		\item Smartphone apps (Android)
		\item Web-apps (Spring MVC)
		\item Desktop applications (Java Swing etc.)
	\end{enumerate}
	\item To develop a consumer smart phone app to demonstrate the working API which automatically ranks a users data according to its relevance
\end{enumerate}

These are the criteria by which the extent of this project's success will be evaluated.

\section{Unique Features}

Many of the aspects of this project have never been seen combined into a single research project before and others (such as query-less context-sensitive scoring of data) have recieved little attention. This project is unique in its endeavour to combine data fusion techniques with context-sensitive scoring in the development of a commercially viable prototype. 
